Using Cell Phone Keyboards is (\NP) Hard

### 2010-06-30 article.tex
Actually fixed the commit

\usepackage{amsmath}
\usepackage{url}
\usepackage{graphicx}
\usepackage{salgorithm}

@@

Typing on a keyboard which has fewer keys than there are letters in the
alphabet can be a painful task.  There are a plethora of input schemes which
attempt to make this task easier, but the one thing they have in common is
that all of these input methods use the standard mapping of numeric keys to
alphabetic numbers of 
$\{2\to\{abc\},
         3\to\{def\}, 4\to\{ghi\}, 5\to\{jkl\}, 6\to\{mno\}, 7\to\{pqrs\},
         8\to\{tuv\}, 9\to\{wxyz\}\}$.
We will consider schemes of rearranging the numbers on the keys to
make messages easier to type.  Most variants of this problem turn out to be
NP-hard, unfortunately.

@@

Here, we consider three different real-world schemes
for $\mathrm{IN}$ (basic typing, T9, and predictive T9), and for each variant
that is proven \NP-hard, we consider the restriction on $P$ that requires that
we keep the alphabet in alphabetical order.  In all cases where it is
computationally feasible we provide results for the case of the English
language, on 8-key keyboards, using the British National Corpus\cite{bnc}
(BNC).  One feature of note is that no matter what scheme we use, the problem
is trivial if the number of keys is not smaller than the number of letters in
the alphabet ($k \ge |A|$).  Using tiny keyboards is only (computationally)
difficult when the keyboards do not have enough keys.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=2in]{phonekeys.jpg}
\end{center}
\caption{A cell phone keyboard}
\label{keypic}
\end{figure}

Initially, we will completely neglect the pauses and concern ourselves solely
with the number of keystrokes required.  This implies that `a' will always
require one button press, `b' will always require two, and so on.  Thus, to get
a baseline of how many keystrokes are required to enter the entirety of our
corpus of words $W$, we count the total number of occurrences of each letter,
and multiply that number of occurrences by the number of button pushes
required by that letter.  Running this on the British National Corpus
using the standard cell phone keyboard layout, we find that entering the
entirety of the 100,106,029 word occurrences in the corpus would require
948,049,046 button presses.

If we examine a cell phone keyboard (Fig.~\ref{keypic}) then we observe some terrible design choices.  The frequently-occurring letters
'r' and 's' require more keystrokes than 'q'!  Surely, a better designed
keyboard can do better than this.  If we just reversed the order of the letters
on the 7 key, from `pqrs' to `srqp', the number of button presses required
would drop to 865,118,331 --- a savings of more than $8.7\%$!  If we assume
that every button press takes an equal amount of time, then this corresponds
the users spending $8.7\%$ less time entering their text messages.  In an
effort to discover how great these savings can be, we define the problem {\sc
BasicCellPhoneTyping} based on the template problem on
page~\pageref{probtemplate}.

@@

By an argument symmetric to the proof of Theorem~\ref{basicthm}, we find that
the optimal layout in this new scheme is to place the letters of a given key in
sorted order, according to frequency.  Thus, the keyboard layout changes to
$2\to[acb], 3\to[edf], 4\to[ihg], 5\to[lkj], 6\to[onm], 7\to[srpq], 8\to[tuv],
9\to[wyxz]$ and requires 678,547,463 button presses to enter the whole corpus.
This layout represents a $28.43\%$ savings, and it has the added benefit that
it does not change which key is mapped to which set of letters.  This makes it
\newword{legacy preserving}, as creating a keyboard with this layout will not
invalidate such advertising gems as 1-800-FLOWERS.  Thus, we can speed up users
by $28.43\%$ (neglecting inter-letter pauses) and not undermine 50 years of
advertising.  This layout represents perhaps the most plausible layout
yet\footnote{The described layout is the only legacy preserving layout in this
paper, and therefore should be considered the most practical suggestion.  It
also has the added benefit of not angering any of the organizations behind the
numbers 1-800-FLOWERS, 1-800-THRIFTY, and 1-800-MARINES}.  After setting up
this baseline for the easy problem, we turn our attention to optimizing
cellphone keyboards for predictive input methods.

@@

Given $G=(V,E)$ and $k$ from {\sc GraphColoring}, we create $$G'=\left(V \cup
\{0,1\}, E \cup \{(0,1)\} \cup \{ (v, 0)~ \forall v \in V\} \right)$$ We then
uniquely number each edge in $E$ with the numbers $1 \ldots |E|$.  For each
edge $e = (u,v)$ numbered $i$, we create the path set 
\begin{eqnarray*}
p_i &=& \{[v,0,b_1(i),b_2(i), b_3(i), \ldots b_{\lceil \log_2 i\rceil}], \\
    & &  ~[u, 0,b_1(i),b_2(i), b_3(i), \ldots b_{\lceil \log_2 i\rceil}]\} \\
\end{eqnarray*}
where $b_1(i)$ is the first digit of $i$ in binary, $b_2(i)$ is the second
digit of $i$ in binary, and so on.  Now we create our path set $$P = \{ [0,1],
[1,0] \} \cup \bigcup_{i = 1}^{|E|} p_i \enspace .$$ We then ask the question
of whether $G'$ and $P$ can be unique-path colored using only $k$ colors.

@@

Because our problem is \NP-complete, we will not try to find a general solution.  In our specific instance, with 26 letters and an eight key keyboard (1 and 0 are reserved for other uses),  we must choose the best layout from among $$\binom{26}{8} \binom{18}{8} \binom{10}{8} = 3,076,291,325,250$$ different possibilities.  Given that, currently, it takes around a tenth of a second to count the button presses required to enter the BNC, an exhaustive search will take 307,629,132,525 seconds, or 9.75 millennia.  Therefore, we have two remaining avenues of attack: find the best answer possible using stochastic methods, or solving a simpler problem.  Using stochastic methods (a simple genetic algorithm), the best layout we found was 
$$\{\{eb\}, \{lcd\}, \{sh\}, \{zmx\}, \{fayg\}, \{toj\}, \{unpv\}, \{irkwq\}\}$$
which required 441,612,049 button presses to enter the entire corpus.   As compared to the requirements of the default layout (443,374,079), this represents a savings of less than 1\%, and is thus not a very attractive alternative to existing layouts.  This low amount of improvement, while not good news for our proposed reordering, is excellent news for all current cell phone users: cell phone keyboard layout seem relatively efficient for T9 text entry!

@@

We left as future work any exploration of the issue of needing to pause between
letters in the basic typing method, and, while there do exist even more
sophisticated input methods\cite{hcimethods}, we have left them unmentioned and unexamined.  Optimizing a keyboard for the basic text input method was relatively straightforward, but, for the T9 input method, optimally using a cell phone keyboard is \NP hard.

\bibliographystyle{abbrv}
\bibliography{bibliography}
\end{document}

>>>>

\usepackage{amsmath}
\usepackage{url}
\usepackage{graphicx}
\usepackage{salgorithm}
\usepackage{verbatim}

@@

Typing on a keyboard which has fewer keys than there are letters in the
alphabet can be a painful task.  There are a plethora of input schemes which
attempt to make this task easier (e.g. \cite{hcimethods,ordering1,godfest} and many more), but the one thing they have in common is
that all of these input methods use the standard mapping of numeric keys to
alphabetic numbers of 
$\{2\to\{abc\}, 3\to\{def\}, 4\to\{ghi\}, 5\to\{jkl\}, 6\to\{mno\}, 7\to\{pqrs\}, 8\to\{tuv\}, 9\to\{wxyz\}\}$.
We will consider schemes of rearranging the numbers on the keys to
make messages easier to type.  Most variants of this problem turn out to be
NP-hard, unfortunately.

@@

We will consider two different real-world schemes for $\mathrm{IN}$ (basic
typing and T9), and for each variant or sub-variant that is proven \NP-hard, we
consider the restriction on $P$ that requires that we keep the alphabet in
alphabetical order.  In all cases where it is computationally feasible we
provide results for the case of the English language, on 8-key keyboards, using
the British National Corpus\cite{bnc} (BNC).  One feature of note is that no
matter what scheme we use, the problem is trivial if the number of keys is not
smaller than the number of letters in the alphabet ($k \ge |A|$).  Using tiny
keyboards is only (computationally) difficult when the keyboards do not have
enough keys.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=2in]{phonekeys.jpg}
\end{center}
\caption{A cell phone keyboard. Each key is mapped to a letter sequence: $2 \to (abc), 3\to (def), 4\to (ghi), 5\to (jkl), 6 \to (mno), 7 \to (pqrs), 8\to (tuv), 9\to (wxyz) $}
\label{keypic}
\end{figure}

We will neglect the pauses and concern ourselves solely with the number of
keystrokes required.  This layout and input method imply that `a' will always
require one button press, `b' will always require two, and so on.  Thus, to get
a baseline of how many keystrokes are required to enter the entirety of our
corpus of words $W$, we count the total number of occurrences of each letter,
and multiply that number of occurrences by the number of button pushes required
by that letter.  Running this on the British National Corpus using the standard
cell phone keyboard layout, we find that entering the entirety of the
100,106,029 word occurrences in the corpus would require 948,049,046 button
presses.

If we examine a cell phone keyboard (Fig.~\ref{keypic}) then we observe some terrible design choices.  The frequently-occurring letters
'r' and 's' require more keystrokes than 'q'!  Surely, a better designed
keyboard can do better than this.  If we just reversed the order of the letters
on the 7 key, from `pqrs' to `srqp', the number of button presses required
would drop to 865,118,331 --- a savings of more than $8.7\%$.  If we assume
that every button press takes an equal amount of time, then this corresponds
the users spending $8.7\%$ less time entering their text messages.  In an
effort to discover how great these savings can be, we define the problem {\sc
BasicCellPhoneTyping} based on the template problem on
page~\pageref{probtemplate}.

@@

By an argument symmetric to the proof of Theorem~\ref{basicthm}, we find that
the optimal layout in this new scheme is to place the letters of a given key in
sorted order, according to frequency.  Thus, the keyboard layout changes to
$2\to[acb], 3\to[edf], 4\to[ihg], 5\to[lkj], 6\to[onm], 7\to[srpq], 8\to[tuv],
9\to[wyxz]$ and requires 678,547,463 button presses to enter the whole corpus.
This layout represents a $28.43\%$ savings, and it has the added benefit that
it does not change which key is mapped to which set of letters.  This makes it
\newword{legacy preserving}, as creating a keyboard with this layout will not
invalidate such advertising gems as 1-800-FLOWERS.  Thus, we can speed up users
by $28.43\%$ (neglecting inter-letter pauses) and not undermine more than half
a century of advertising.  This layout represents perhaps the most plausible
layout yet\footnote{The described layout is the only legacy preserving layout
in this paper, and therefore should be considered the most practical
suggestion.  It also has the added benefit of not angering any of the
organizations behind the numbers 1-800-FLOWERS, 1-800-THRIFTY, and
1-800-MARINES}.  After setting up this baseline for the easy problem, we turn
our attention to optimizing cellphone keyboards for predictive input methods.

@@

Given $G=(V,E)$ and $k$ from {\sc GraphColoring}, we create $$G'=\left(V \cup
\{0,1\}, E \cup \{(0,1), (1,0), (0,0), (1,1)\} \cup \{ (v, 0)~ \forall v \in V\} \right)$$ We then
uniquely number each edge in $E$ with the numbers $1 \ldots |E|$.  For each
edge $e = (u,v)$ numbered $i$, we create the path set 
\begin{eqnarray*}
p_i &=& \{[v,0,b_1(i),b_2(i), b_3(i), \ldots b_{\lceil \log_2 i\rceil}], \\
    & &  ~[u, 0,b_1(i),b_2(i), b_3(i), \ldots b_{\lceil \log_2 i\rceil}]\} \\
\end{eqnarray*}
where $b_1(i)$ is the first digit of $i$ in binary, $b_2(i)$ is the second
digit of $i$ in binary, and so on.  Now we create our path set $$P = \{ [0,1],
[1,0] \} \cup \bigcup_{i = 1}^{|E|} p_i \enspace .$$ We then ask the question
of whether $G'$ and $P$ can be unique-path colored using only $k$ colors.

@@

Because our problem is \NP-complete, we will not try to find a general solution.  In our specific instance, with 26 letters and an eight key keyboard (1 and 0 are reserved for other uses),  we must choose the best layout from among $$\binom{26}{8} \binom{18}{8} \binom{10}{8} = 3,076,291,325,250$$ different possibilities.  Given that, currently, it takes around a tenth of a second to count the button presses required to enter the BNC, an exhaustive search will take 307,629,132,525 seconds, or 9.75 millennia.  Therefore, we have two remaining avenues of attack: find the best answer possible using stochastic methods, or solving a simpler problem.  Using stochastic methods (a simple genetic algorithm), the best layout we found was 
$$\{\{eb\}, \{lcd\}, \{sh\}, \{zmx\}, \{fayg\}, \{toj\}, \{unpv\}, \{irkwq\}\}$$
which required 441,612,049 button presses to enter the entire corpus.  Our genetic algorithm was quite simple: From an initial gene pool of random layouts, we discarded all but the best layouts (the survivors).  Then, we added to the gene pool random mutations of the survivors to fill out the gene pool back to its original size, and repeated the process of selection and mutation for some time.  As compared to the requirements of the default layout (443,374,079), this represents a savings of less than 1\%, and is thus not a very attractive alternative to existing layouts.  This low amount of improvement, while not good news for our proposed reordering, is excellent news for all current cell phone users: the default cell phone keyboard layout seems relatively efficient for T9 text entry in English!

@@

We left as future work any exploration of the issue of needing to pause between
letters in the basic typing method, and, while there do exist even more
sophisticated input methods\cite{hcimethods}, we have left them unmentioned and unexamined.  Optimizing a keyboard for the basic text input method was relatively straightforward, but, for the T9 input method, optimally using a cell phone keyboard is \NP\ hard.

\bibliographystyle{abbrv}
\bibliography{bibliography}

\newpage
\appendix
\section{A Genetic Algorithm for Discovering Efficient T9 Layouts}
{\tiny
\verbatiminput{genetic.cpp}
}

\section{Code for Analyzing All Alphabetic-Order T9 Layouts}
{\tiny
\verbatiminput{notrcount.cpp}
}
\end{document}
